---
title: "In Class Assignment 10 - Building a Research Cohort"
subtitle: "Relational Databases and SQL Programming | Spring 2026"
author: "Ruby He"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Load Packages
```{r}
library(CDMConnector)
library(DBI)
library(duckdb)

library(tidyverse)
library(gtsummary)
theme_gtsummary_compact()
library(readxl)
library(ggplot2)
library(survival)
library(ggsurvfit)
library(labelled)
```

### Connect to Database

```{r}
covid_df = DBI::dbConnect(duckdb::duckdb(),
                          eunomiaDir("synthea-covid19-200k"))

dbListTables(covid_df)
```

### Step 1 - Patient Demographics

*For all patients in the database, retrieve their demographic information including birth date, gender, race, and ethnicity.*

```{r}
demo = dbGetQuery(covid_df, 
                  "SELECT person_id, 
                    birth_datetime, 
                    gender_source_value AS gender, 
                    race_source_value AS race,
                    ethnicity_source_value AS ethnicity
                    FROM person;")

demo %>% 
  head() %>% 
  rmarkdown::paged_table()
```

### Step 2 - COVID-19 Inclusion Criteria

*Identify all patients with a diagnosis of COVID-19. For each patient identify their first diagnosis date*

* domain_id = 'Condition'
* concept_name = 'COVID-19'

```{r}
covid_entry = dbGetQuery(covid_df, 
                        "SELECT c.person_id, 
                        c_condition_occurrence.concept_name AS condition_name,
                        MIN(c.condition_start_date) AS first_covid_dx
                        FROM condition_occurrence AS c
                        INNER JOIN concept AS c_condition_occurrence
                          ON c.condition_concept_id = c_condition_occurrence.concept_id
                        WHERE c_condition_occurrence.domain_id = 'Condition' AND 
                          c_condition_occurrence.concept_name = 'COVID-19'
                        GROUP BY c.person_id, condition_name;")

covid_entry %>% 
  head() %>% 
  rmarkdown::paged_table()
```

### Step 3 - Death Outcome 

*Identify all patients who have died. For each patient identify their death date.*

```{r}
death_outcome = dbGetQuery(covid_df, 
                          "SELECT person_id, death_date
                          FROM death;")

death_outcome %>% 
  head() %>% 
  rmarkdown::paged_table()
```


### Final Cohort 

```{r}
cohort = dbGetQuery(covid_df,
                         
  "WITH demographics AS 
      (
        SELECT person_id, birth_datetime, 
        gender_source_value AS gender, 
        race_source_value AS race, 
        ethnicity_source_value AS ethnicity
        FROM person
      ),
    covid_dx AS 
      (
        SELECT c.person_id, 
          c_condition_occurrence.concept_name AS condition_name,
          MIN(c.condition_start_date) AS first_covid_dx
        FROM condition_occurrence AS c
        INNER JOIN concept AS c_condition_occurrence
          ON c.condition_concept_id = c_condition_occurrence.concept_id
        WHERE c_condition_occurrence.domain_id = 'Condition' AND
              c_condition_occurrence.concept_name = 'COVID-19'
        GROUP BY c.person_id, condition_name
      ),
    death_dates AS 
      (
        SELECT person_id, death_date
        FROM death
      )
    
      
    SELECT demo.person_id, demo.gender, demo.race, demo.ethnicity,
    DATEDIFF('year', demo.birth_datetime, covid.first_covid_dx) AS age_at_covid_dx,
    covid.first_covid_dx, death.death_date,
    DATEDIFF('day', covid.first_covid_dx, death.death_date) AS days_covid_to_death,
    CASE WHEN death.death_date IS NOT NULL THEN 1 
              ELSE 0 END AS death_yn
    FROM demographics AS demo
    INNER JOIN covid_dx AS covid
      ON demo.person_id = covid.person_id
    LEFT JOIN death_dates AS death
      ON covid.person_id = death.person_id
    WHERE DATEDIFF('year', demo.birth_datetime, covid.first_covid_dx) >= 65
     AND ((death.death_date >= covid.first_covid_dx) OR (death.death_date IS NULL));")

cohort %>% 
  head() %>% 
  rmarkdown::paged_table()
```

### Cohort Summary Table 

```{r}
cohort = cohort %>% 
  mutate(
    gender = case_when(
      gender == "M" ~ "Male",
      gender == "F" ~ "Female",
      TRUE ~ gender),
    race = str_to_title(race),
    ethnicity = str_to_title(ethnicity)) %>% 
  set_variable_labels(
    death_yn = "Death",
    age_at_covid_dx = "Age",
    gender = "Gender",
    race = "Race",
    ethnicity = "Ethnicity",
    days_covid_to_death = "Time to Death")

cohort %>% 
  select(age_at_covid_dx, gender, race, ethnicity,
         death_yn, days_covid_to_death) %>% 
  tbl_summary(
    by = death_yn,
    type = list(age_at_covid_dx    ~ "continuous2",
                days_covid_to_death ~ "continuous2"),
    statistic = list(all_continuous2() ~ c("{mean} ({sd})",
                                           "{median} ({p25}, {p75})",
                                           "{min}, {max}"))) %>% 
  bold_labels() 
```

### Survival Plot

```{r}
cohort %>% 
  mutate(
    time_30  = pmin(days_covid_to_death, 30, na.rm = TRUE),
    event_30 = ifelse(death_yn == 1 & days_covid_to_death <= 30, 1, 0)) %>% 
  survfit2(Surv(time_30, event_30) ~ gender, data = .) %>%
  ggsurvfit() +
  add_confidence_interval() 
```



