---
title: "Session 10 In Class Practice Session - OMOP Exploration"
subtitle: "Relational Databases and SQL Programming | Spring 2026"
author: "Zoe Verzani"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Load Packages
```{r}
library(CDMConnector)
library(DBI)
library(duckdb)
library(tidyverse)
library(gtsummary)
theme_gtsummary_compact()
library(readxl)
library(ggplot2)
library(survival)
library(ggsurvfit)
library(labelled)

# usethis::edit_r_environ()
# add this - EUNOMIA_DATA_FOLDER="my/data/path" - SAVE IT 
# Restart R — Session → Restart R in RStudio
```

## Synthea Databases 

We will be using a dataset provided by [Synthea](https://synthetichealth.github.io/synthea/) — an open-source, synthetic patient generator that models the medical history of synthetic patients. Synthea produces realistic but entirely fictional patient records, covering demographics, encounters, diagnoses, medications, procedures, and more, without any risk of exposing real patient data.

The dataset has been mapped to the OMOP Common Data Model [(CDM)](https://ohdsi.github.io/CommonDataModel/cdm53.html), a standardized structure maintained by the OHDSI community that enables consistent analysis across different healthcare databases. This allows us to apply the same analytical code regardless of the underlying data source.

```{r}
exampleDatasets()
```

## COVID-19 Database 

```{r}
covid_df = DBI::dbConnect(duckdb::duckdb(),
                          eunomiaDir("synthea-covid19-200k"))

#dbListTables(covid_df)
```

## Part 1 - Data Exploration

### Persons Table 

The person table is the foundation of every OMOP database, containing one row per unique patient. It stores core demographic attributes like birth date, gender, race, and ethnicity.

```{r patient table head}
dbGetQuery(covid_df, 
           "SELECT *
           FROM person
           LIMIT 100;") %>% 
  rmarkdown::paged_table()
```


#### Question 1) 

How many patients are in the database?

```{r q1 patient count}
query1 = dbGetQuery(covid_df, 
                    "SELECT COUNT(*) AS pat_count
                    FROM person;")

query1
```

#### Question 2)

Create a demographics table displaying the birth date, gender, race, and ethnicity of each patient,

```{r q2 demo query}
query2 = dbGetQuery(covid_df, 
                    "SELECT person_id, birth_datetime, 
                    gender_source_value AS gender, race_source_value AS race,
                    ethnicity_source_value AS ethnicity
                    FROM person;")
query2 %>% 
  head() %>% 
  rmarkdown::paged_table()
```

```{r demo table}
demo_table = query2 %>% 
  mutate(
    gender = case_when(
    gender == "F"  ~ "Female",
    gender == "M" ~ "Male",
    TRUE ~ gender),
    race = str_to_title(race),
    ethnicity = str_to_title(ethnicity)) %>%
tbl_summary(
  include = -person_id,
  label = list(
    gender = "Gender",
    birth_datetime = "Birthdate",
    race = "Race",
    ethnicity = "Ethnicity")) %>% 
  bold_labels()

demo_table
```

### Concept Table 

The concept table is the vocabulary backbone of the entire OMOP Common Data Model. Rather than storing clinical terms as free text (e.g., "Inpatient Visit", "COVID-19"), OMOP stores integer IDs — called concept IDs — in every clinical table. The concept table is what translates those integers into meaningful information.

```{r concept table head}
dbGetQuery(covid_df, 
          "SELECT *
          FROM concept
          LIMIT 100;") %>% 
  rmarkdown::paged_table()
```

Key columns in the concept table:

* concept_id — the unique integer identifier (this is what clinical tables store)
* concept_name — the human-readable label (e.g., "Type 2 diabetes mellitus")
* domain_id — what kind of thing it is: "Condition", "Drug", "Visit", "Measurement", etc.

```{r domains}
domains = dbGetQuery(covid_df, 
          "SELECT DISTINCT(domain_id) AS domain_id
           FROM concept;")

as.list(domains)
```
* vocabulary_id — which coding system it came from (e.g., SNOMED, RxNorm, ICD10CM)

```{r vocabulary}
vocabulary = dbGetQuery(covid_df, 
          "SELECT DISTINCT(vocabulary_id) AS vocabulary_id
           FROM concept;")

as.list(vocabulary)
```
* concept_class_id — a finer classification within the vocabulary
* standard_concept — whether this concept is the standardized version ("S") or a source mapping

### Death Table 

The death table records the date and optional cause of death for patients who have died, with at most one row per person. 

```{r death table head}
dbGetQuery(covid_df, 
          "SELECT *
          FROM death
          LIMIT 100;") %>% 
  rmarkdown::paged_table()
```

#### Question 3) 

How many patients died in the database?

```{r q3 death count}
query3 = dbGetQuery(covid_df, 
                    "SELECT COUNT(*) AS per_count
                    FROM death")

query3
```

#### Question 4)

What were the most common causes of death among deceased patients?

```{r q4 cause of death}
query4 = dbGetQuery(covid_df, 
                    "SELECT d.person_id, d.cause_concept_id, c.concept_name AS cause_of_death
                    FROM death AS d
                    INNER JOIN concept AS c
                      ON d.cause_concept_id = c.concept_id;") 

query4 %>% 
  head() %>% 
  rmarkdown::paged_table()
```

```{r cause of death table}
death_table = query4 %>% 
  tbl_summary(
    include = -c(person_id, cause_concept_id),
    label = list(
      cause_of_death = "Cause of Death"),
    sort = list(cause_of_death ~ "frequency")) %>% 
    bold_labels()

death_table
```

### Visit Occurrence Table

The visit occurrence table captures every encounter a patient has had with the healthcare system, such as inpatient admissions, outpatient appointments, and emergency visits. Each row represents a single visit with a start and end date, a visit type.

```{r visit table head}
dbGetQuery(covid_df, 
           "SELECT *
           FROM visit_occurrence
           LIMIT 10;") %>% 
  rmarkdown::paged_table()
```

#### Question 5)

How many encounter (visit_count) took place for each visit type, how many unique patients (patient_count) have had each visit type, and what is the average visit length (avg_visit_length), minimum and maximum visit length, for each visit type?


```{r q5 visit types}
query5 = dbGetQuery(covid_df, 
                    "SELECT c_visit_type.concept_name AS visit_type,
                    COUNT(v.visit_occurrence_id) AS visit_count, 
                    COUNT(DISTINCT v.person_id) AS patient_count,
                    AVG(DATEDIFF('day', v.visit_start_date, v.visit_end_date)) AS avg_visit_length,
                    MIN(DATEDIFF('day', v.visit_start_date, v.visit_end_date)) AS min_visit_length, 
                    MAX(DATEDIFF('day', v.visit_start_date, v.visit_end_date)) AS max_visit_length
                    FROM visit_occurrence AS v
                    INNER JOIN concept AS c_visit_type
                      ON v.visit_concept_id = c_visit_type.concept_id
                    GROUP BY visit_type;")


rmarkdown::paged_table(query5)
```

### Drug Exposure 

The drug exposure table captures every instance of a patient being exposed to a drug — this includes prescriptions written, prescriptions dispensed, and drugs administered directly by a provider. Each row represents one drug exposure event for one patient.

```{r drug table head}
dbGetQuery(covid_df, 
           "SELECT *
           FROM drug_exposure
           LIMIT 10;") %>% 
  rmarkdown::paged_table()
```

#### Question 6) 

How many patients have a recorded Influenza vaccine? 

* concept_name = 'Influenza, seasonal, injectable, preservative free'
* domain_id = 'Drug'

```{r q6 influenza}
query6 = dbGetQuery(covid_df,
           "SELECT COUNT(DISTINCT d.person_id) AS person_count
           FROM drug_exposure AS d
           INNER JOIN concept AS c_drug
            ON d.drug_concept_id = c_drug.concept_id
          WHERE c_drug.concept_name = 'Influenza, seasonal, injectable, preservative free' AND
            c_drug.domain_id = 'Drug';")

query6 
```


### Condition Occurence

The condition occurrence table records every diagnosis, sign, or symptom that has been documented for a patient — whether observed by a provider, coded from a clinical note, or reported by the patient themselves. It is one of the most frequently used tables in OMOP and serves as the primary source for defining inclusion criteria, identifying comorbidities, and capturing outcomes in observational research.


```{r condition table head}
dbGetQuery(covid_df, 
           "SELECT *
           FROM condition_occurrence;") %>% 
  rmarkdown::paged_table()
```

#### Question 7)

At any given visit, a patient can have multiple conditions recorded simultaneously. For example, during a single inpatient admission a patient might have a headache, a history of diabetes, and a COVID-19 infection all coded as separate condition entries linked to the same visit_occurrence_id.

Using the condition_occurrence table, count the number of conditions recorded at each visit. Then use DENSE_RANK() to rank each visit from the most conditions recorded to the least.

```{r q7 condition rank}
query7 = dbGetQuery(covid_df, 
                    "
                    WITH visit_counts AS
                      (SELECT visit_occurrence_id, person_id, 
                      COUNT(condition_occurrence_id) AS condition_count
                      FROM condition_occurrence
                      GROUP BY visit_occurrence_id, person_id
                      ORDER BY condition_count DESC)
                      
                      SELECT visit_occurrence_id, person_id, condition_count,
                      DENSE_RANK() OVER(ORDER BY condition_count DESC) AS visit_rank
                      FROM visit_counts;
                    ")

query7 %>% 
  head() %>% 
  rmarkdown::paged_table()
```

#### Question 8)

8) What were the conditions recorded at the 1st ranked visits. You can hard code the visit_occurrence_id.

```{r q8 explore conditions}
query8 = dbGetQuery(covid_df, 
                   "SELECT visit_occurrence_id, c.person_id, c.condition_concept_id,
                   c_condition.concept_name AS diagnosis
                   FROM condition_occurrence AS c
                   INNER JOIN concept AS c_condition
                    ON c.condition_concept_id = c_condition.concept_id
                   WHERE visit_occurrence_id = 1304452;")
                   

query8 %>% 
  rmarkdown::paged_table()
```

